---
title: "Demo 03: Simple visuals for high-dimensional data"
format: html
---

# More fun with penguins

**The graphs below don't have proper titles, axis labels, legends, etc.  Please take care to do this on your own graphs.** Throughout this demo we will use the [`palmerpenguins` dataset](https://allisonhorst.github.io/palmerpenguins/articles/intro.html). To access the data, you will need to install the `palmerpenguins` package:

```{r}
#| eval: false
install.packages("palmerpenguins")
```

We load the `penguins` data in the same way as the previous demos:

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(palmerpenguins)
data(penguins)
head(penguins)
```

# Pairs plot with [`GGally`](https://ggobi.github.io/ggally/index.html)

We will use the [`GGally`](https://ggobi.github.io/ggally/index.html) package to make pairs plots in `R` with `ggplot` figures. You need to install the package:

```{r}
#| eval: false
install.packages("GGally")
```

Next, we'll load the package and create a pairs plot of just the continuous variables using `ggpairs`. The main arguments you have to worry about for `ggpairs` are `data`, `columns`, and `mapping`:

+ `data`: specifies the dataset

+ `columns`: Columns of data you want in the plot (can specify with vector of column names or numbers referring to the column indices)
+ `mapping`: aesthetics using `aes()`. Most important one is `aes(color = <variable name>)`

First, let’s create a pairs plot by specifying `columns` as the four columns of continuous variables (columns 3 through 6):

```{r}
#| warning: false
#| message: false
library(GGally)
penguins |> ggpairs(columns = 3:6)
```

Obviously this suffers from over-plotting so we'll want to adjust the `alpha`. An annoying thing is that we specify the `alpha` directionly with `aes` when using `ggpairs`:

```{r}
#| warning: false
#| message: false
penguins |> ggpairs(columns = 3:6, mapping = aes(alpha = 0.5))
```

Plots along the diagonal show marginal distributions. Plots along the off-diagonal show joint (pairwise) distributions or statistical summaries (e.g., correlation) to avoid redundancy. The matrix of plots is symmetric; e.g., entry (1,2) shows the same distribution as entry (2,1). However, entry (1,2) and entry (2,1) display different bits of information (or alternative plots) about the same distribution.

We could also specify categorical variables in the plot. We also don't need to specify column indices if we just `select` which columns to use beforehand:

```{r}
#| warning: false
#| message: false
penguins |> 
  dplyr::select(bill_length_mm, body_mass_g, species, island) |>
  ggpairs(mapping = aes(alpha = 0.5))
```

Alternatively, we can use the mapping argument to display these categorical variables in a different manner - and arguably more efficiently:

```{r}
#| warning: false
#| message: false
penguins |> 
  ggpairs(columns = c("bill_length_mm", "body_mass_g", "island"),
          mapping = aes(alpha = 0.5, color = species))
```

The `ggpairs` function in `GGally` is very flexible and customizable with regards to which figures are displayed in the various panels. [I encourage you to check out the vignettes and demos on the package website for more examples.](https://ggobi.github.io/ggally/articles/ggpairs.html) For instance, in the pairs plot below I decide to display the regression lines and make other adjustments to the off-diagonal figures:

```{r}
#| warning: false
#| message: false
penguins |> 
  ggpairs(columns = c("bill_length_mm", "body_mass_g", "island"),
          mapping = aes(alpha = 0.5, color = species), 
          lower = list(
            continuous = "smooth_lm", 
            combo = "facetdensitystrip"
          ),
          upper = list(
            continuous = "cor",
            combo = "facethist"
          )
  )
```

You can also proceed to customize the pairs plot in the same manner as `ggplot` figures:

```{r}
#| warning: false
#| message: false
penguins |>
  dplyr::select(species, body_mass_g, ends_with("_mm")) |>
  ggpairs(mapping = aes(color = species, alpha = 0.5),
          columns = c("flipper_length_mm", "body_mass_g",
                      "bill_length_mm", "bill_depth_mm")) +
  scale_colour_manual(values = c("darkorange","purple","cyan4")) +
  scale_fill_manual(values = c("darkorange","purple","cyan4")) +
  theme_bw() +
  theme(strip.text = element_text(size = 7))
```


# Correlograms with [`ggcorrplot`](https://rpkgs.datanovia.com/ggcorrplot/) 

We can visualize the correlation matrix for the variables in a dataset using the [`ggcorrplot`](https://rpkgs.datanovia.com/ggcorrplot/) package. You need to install the package:

```{r}
#| eval: false
install.packages("ggcorrplot")
```

Next, we'll load the package and create a __correlogram__ using only the continuous variables. To do this, we first need to compute the __correlation matrix__ for these variables:

```{r}
penguins_cor_matrix <- penguins |>
  dplyr::select(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) |>
  cor(use = "complete.obs")
penguins_cor_matrix
```

__NOTE__: Since there are missing values in the `penguins` data we need to indicate in the `cor()` function how to handle missing values using the `use` argument. By default, the correlations are returned as `NA`, which is not what we want. Instead, we can change this to only use observations without `NA` values for the considered columns (see `help(cor)` for more options).

Now, we can create the correlogram using `ggcorrplot()` using this correlation matrix:

```{r}
library(ggcorrplot)
ggcorrplot(penguins_cor_matrix)
```

There are several ways we can improve this correlogram:

- we can avoid redundancy by only using one half of matrix by changing the `type` input: the default is `full`, we can make it `lower` or `upper` instead:

```{r}
ggcorrplot(penguins_cor_matrix, type = "lower")
```

- we can rearrange the variables using hierarchical clustering so that variables displaying stronger levels of correlation are closer together along the diagonal by setting `hc.order = TRUE`:

```{r}
ggcorrplot(penguins_cor_matrix, type = "lower", hc.order = TRUE)
```

- if we want to add the correlation values directly to the plot, we can include those labels setting `lab = TRUE` - but we should round the correlation values first using the `round()` function:

```{r}
ggcorrplot(round(penguins_cor_matrix, digits = 4), 
           type = "lower", hc.order = TRUE, lab = TRUE)
```

- if we want to place more stress on the correlation magnitude, we can change the `method` input to `circle` so that the size of the displayed circles is mapped to the absolute value of the correlation value:

```{r}
#| warning: false
#| message: false
ggcorrplot(penguins_cor_matrix, type = "lower", hc.order = TRUE,
           method = "circle")
```

You can ignore the `Warning` message that is displayed - just from the differences in `ggplot` implementation.

# Heatmaps to display dataset structure with color

Heatmaps provide a way to display structure of the dataset using the fill of tiles in a matrix. The fill of the tiles is mapped to a variable's standardized value (i.e., (x - mean(x)) / sd(x)). There is a convenient function in `R` called `heatmap` to create this type of figure:

```{r}
heatmap(as.matrix(dplyr::select(penguins, 
                                bill_length_mm, bill_depth_mm, 
                                flipper_length_mm, body_mass_g)),
        scale = "column", 
        Rowv = NA, Colv = NA)
```

In order to manually create this figure, we'll need to pivot our dataset from wide to long using the `pivot_longer()` function. This results in a dataset with one row per observation and variable combination. Then we use `geom_tile` as the geometric object with the standardized value mapped to the fill:

```{r}
penguins |>
  mutate(penguin_index = as.factor(paste0("Penguin-", 1:n()))) |>
  dplyr::select(penguin_index, bill_length_mm, bill_depth_mm, 
                flipper_length_mm, body_mass_g) |>
  pivot_longer(bill_length_mm:body_mass_g,
               names_to = "variable",
               values_to = "raw_value") |>
  group_by(variable) |>
  mutate(std_value = (raw_value - mean(raw_value, na.rm = TRUE)) / 
           sd(raw_value, na.rm = TRUE)) |>
  ungroup() |>
  ggplot(aes(x = variable, y = penguin_index, fill = std_value)) +
  geom_tile() +
  theme_light() +
  theme(legend.position = "bottom",
        axis.text.y = element_text(size = 2)) 
```

In order to provide some notion of the correlation structure between variables, it's useful to reorder the observations in the heatmap display by some variable:

```{r}
penguins |>
  mutate(penguin_index = as.factor(paste0("Penguin-", 1:n())),
         penguin_index = fct_reorder(penguin_index, body_mass_g,
                                     # Ignore the missings when reordering
                                     .na_rm = TRUE)) |>
  dplyr::select(penguin_index, bill_length_mm, bill_depth_mm, 
                flipper_length_mm, body_mass_g) |>
  pivot_longer(bill_length_mm:body_mass_g,
               names_to = "variable",
               values_to = "raw_value") |>
  group_by(variable) |>
  mutate(std_value = (raw_value - mean(raw_value, na.rm = TRUE)) / 
           sd(raw_value, na.rm = TRUE)) |>
  ungroup() |>
  ggplot(aes(x = variable, y = penguin_index, fill = std_value)) +
  geom_tile() +
  scale_fill_gradient(low = "darkblue", high = "darkorange") +
  theme_light() +
  theme(legend.position = "bottom",
        axis.text.y = element_text(size = 2)) 
```


# Parallel coordinates plot with [`GGally`](https://ggobi.github.io/ggally/index.html)

In a __parallel coordinates__ plot, we create an axis for each varaible and align these axes side-by-side, drawing lines between observations from one axis to the next. This can be useful for visualizing structure among __both__ the variables and observations in our dataset. These are useful when working with a moderate number of observations and variables - but can be overwhelming with too many.

We use the `ggparcoord()` function from the [`GGally`](https://ggobi.github.io/ggally/index.html) package to make parallel coordinates plots:

```{r}
penguins |>
  ggparcoord(columns = 3:6)
```

There are several ways we can modify this parallel coordinates plot:

- we should __always__ adjust the transparency of the lines using the `alphaLines` input to help handle overlap:

```{r}
penguins |>
  ggparcoord(columns = 3:6, alphaLines = .2)
```

- we can color each observation's lines by a categorical variable, which can be useful for revealing group structure:

```{r}
penguins |>
  ggparcoord(columns = 3:6, alphaLines = .2, groupColumn = "species")
```

- we can change how the y-axis is constructed by modifying the `scale` input, which by default is `std` that is simply subtracting the mean and dividing by the standard deviation. We could instead use `uniminmax` so that minimum of the variable is zero and the maximum is one:

```{r}
penguins |>
  ggparcoord(columns = 3:6, alphaLines = .2, groupColumn = "species",
             scale = "uniminmax")
```

- we can also reorder the variables a number of different ways with the `order` input (see `help(ggparcoord)` for details). There appears to be some weird errors however with the different options, but you can still manually provide the order of indices as follows:

```{r}
penguins |>
  ggparcoord(columns = 3:6, alphaLines = .2, groupColumn = "species",
             order = c(6, 5, 3, 4))
```
